---
track: Data Science
course: Getting and Cleaning Data
title: codebook.Rmd
date: January 22, 2016
output: html_document
---

<hr>

This is an R markdown document describing all the variables included in the tidy data sets created as the result of the course project, as well as the processing steps used to create the tidy data sets.

<hr>

## Variables

- **subject**
    
    Type: integer<br/>
    Description: subject ID<br/>
    Values: 1, 2, 3, 4, 5, 6

- **activity**

    Type: character<br/>
    Description: activity type<br/>
    Values: "WALKING", "WALKING_UPSTAIRS", "WALKING_DOWNSTAIRS", "SITTING", "STANDING", "LAYING"

- **timed_body_acceleration_mean_angle_x**

    Type: numeric<br/>
    Description: mean of timed body acceleration - angle X
    
- **timed_body_acceleration_mean_angle_y**

    Type: numeric<br/>
    Description: mean of timed body acceleration - angle Y

- **timed_body_acceleration_mean_angle_z**

    Type: numeric<br/>
    Description: mean of timed body acceleration - angle Z

- **timed_body_acceleration_standard_deviation_angle_x**

    Type: numeric<br/>
    Description: Standard deviation of timed body acceleration - angle X
    
- **timed_body_acceleration_standard_deviation_angle_y**

    Type: numeric<br/>
    Description: Standard deviation of timed body acceleration - angle Y

- **timed_body_acceleration_standard_deviation_angle_z**

    Type: numeric<br/>
    Description: Standard deviation of timed body acceleration - angle Z

- **timed_gravity_acceleration_mean_angle_x**

    Type: numeric<br/>
    Description: Mean of timed gravity acceleration - angle X

- **timed_gravity_acceleration_mean_angle_y**

    Type: numeric<br/>
    Description: Mean of timed gravity acceleration - angle Y

- **timed_gravity_acceleration_mean_angle_z**

    Type: numeric<br/>
    Description: Mean of timed gravity acceleration - angle Z

- **timed_gravity_acceleration_standard_deviation_angle_x**

    Type: numeric<br/>
    Description: Standard deviation of timed gravity acceleration - angle X

- **timed_gravity_acceleration_standard_deviation_angle_y**

    Type: numeric<br/>
    Description: Standard deviation of timed gravity acceleration - angle Y

- **timed_gravity_acceleration_standard_deviation_angle_z**

    Type: numeric<br/>
    Description: Standard deviation of timed gravity acceleration - angle Z

- **timed_body_acceleration_jerk_signal_mean_angle_x**

    Type: numeric<br/>
    Description: Mean of timed body acceleration jerk signal - angle X

- **timed_body_acceleration_jerk_signal_mean_angle_y**

    Type: numeric<br/>
    Description: Mean of timed body acceleration jerk signal - angle Y

- **timed_body_acceleration_jerk_signal_mean_angle_z**

    Type: numeric<br/>
    Description: Mean of timed body acceleration jerk signal - angle Z

- **timed_body_acceleration_jerk_signal_standard_deviation_angle_x**

    Type: numeric<br/>
    Description: Standard deviation of timed body acceleration jerk signal - angle X

- **timed_body_acceleration_jerk_signal_standard_deviation_angle_y**

    Type: numeric<br/>
    Description: Standard deviation of timed body acceleration jerk signal - angle Y

- **timed_body_acceleration_jerk_signal_standard_deviation_angle_z**

    Type: numeric<br/>
    Description: Standard deviation of timed body acceleration jerk signal - angle Z

- **timed_body_gyroscope_mean_angle_x**

    Type: numeric<br/>
    Description: Mean of timed body gyroscope - angle X

- **timed_body_gyroscope_mean_angle_y**

    Type: numeric<br/>
    Description: Mean of timed body gyroscope - angle Y

- **timed_body_gyroscope_mean_angle_z**

    Type: numeric<br/>
    Description: Mean of timed body gyroscope - angle Z

- **timed_body_gyroscope_standard_deviation_angle_x**

    Type: numeric<br/>
    Description: Standard deviation of timed body gyroscope - angle X
    
- **timed_body_gyroscope_standard_deviation_angle_y**

    Type: numeric<br/>
    Description: Standard deviation of timed body gyroscope - angle Y

- **timed_body_gyroscope_standard_deviation_angle_z**

    Type: numeric<br/>
    Description: Standard deviation of timed body gyroscope - angle Z

- **timed_body_gyroscope_jerk_signal_mean_angle_x**

    Type: numeric<br/>
    Description: Mean of timed body gyroscope jerk signal - angle X

- **timed_body_gyroscope_jerk_signal_mean_angle_y**

    Type: numeric<br/>
    Description: Mean of timed body gyroscope jerk signal - angle Y

- **timed_body_gyroscope_jerk_signal_mean_angle_z**

    Type: numeric<br/>
    Description: Mean of timed body gyroscope jerk signal - angle Z

- **timed_body_gyroscope_jerk_signal_standard_deviation_angle_x**

    Type: numeric<br/>
    Description: Standard deviation of timed body gyroscope jerk signal - angle X

- **timed_body_gyroscope_jerk_signal_standard_deviation_angle_y**

    Type: numeric<br/>
    Description: Standard deviation of timed body gyroscope jerk signal - angle Y

- **timed_body_gyroscope_jerk_signal_standard_deviation_angle_z**

    Type: numeric<br/>
    Description: Standard deviation of timed body gyroscope jerk signal - angle Z

- **timed_body_acceleration_magnitude_mean**

    Type: numeric<br/>
    Description: Mean of timed body acceleration magnitude

- **timed_body_acceleration_magnitude_standard_deviation**

    Type: numeric<br/>
    Description: Standard deviation of timed body acceleration magnitude

- **timed_gravity_acceleration_magnitude_mean**

    Type: numeric<br/>
    Description: Mean of timed gravity acceleration magnitude

- **timed_gravity_acceleration_magnitude_standard_deviation**

    Type: numeric<br/>
    Description: Standard deviation of timed gravity acceleration magnitude

- **timed_body_acceleration_jerk_signal_magnitude_mean**

    Type: numeric<br/>
    Description: Mean of timed body acceleration jerk signal magnitude

- **timed_body_acceleration_jerk_signal_magnitude_standard_deviation**

    Type: numeric<br/>
    Description: Standard deviation of timed body acceleration jerk signal magnitude

- **timed_body_gyroscope_magnitude_mean**

    Type: numeric<br/>
    Description: Mean of timed body gyroscope magnitude

- **timed_body_gyroscope_magnitude_standard_deviation**

    Type: numeric<br/>
    Description: Standard deviation of timed body gyroscope magnitude

- **timed_body_gyroscope_jerk_signal_magnitude_mean**

    Type: numeric<br/>
    Description: Mean of timed body gyroscope jerk signal magnitude

- **timed_body_gyroscope_jerk_signal_magnitude_standard_deviation**

    Type: numeric<br/>
    Description: Standard deviation of timed body gyroscope jerk signal magnitude

- **fast_fourier_transform_body_acceleration_mean_angle_x**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body acceleration - angle X

- **fast_fourier_transform_body_acceleration_mean_angle_y**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body acceleration - angle Y

- **fast_fourier_transform_body_acceleration_mean_angle_z**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body acceleration - angle Z

- **fast_fourier_transform_body_acceleration_standard_deviation_angle_x**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body acceleration - angle X

- **fast_fourier_transform_body_acceleration_standard_deviation_angle_y**
    
    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body acceleration - angle Y

- **fast_fourier_transform_body_acceleration_standard_deviation_angle_z**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body acceleration - angle Z

- **fast_fourier_transform_body_acceleration_jerk_signal_mean_angle_x**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body acceleration jerk signal - angle X

- **fast_fourier_transform_body_acceleration_jerk_signal_mean_angle_y**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body acceleration jerk signal - angle Y

- **fast_fourier_transform_body_acceleration_jerk_signal_mean_angle_z**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body acceleration jerk signal - angle Z

- **fast_fourier_transform_body_acceleration_jerk_signal_standard_deviation_angle_x**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body acceleration jerk signal - angle X

- **fast_fourier_transform_body_acceleration_jerk_signal_standard_deviation_angle_y**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body acceleration jerk signal - angle Y

- **fast_fourier_transform_body_acceleration_jerk_signal_standard_deviation_angle_z**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body acceleration jerk signal - angle Z

- **fast_fourier_transform_body_gyroscope_mean_angle_x**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body gyroscope - angle X

- **fast_fourier_transform_body_gyroscope_mean_angle_y**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body gyroscope - angle Y

- **fast_fourier_transform_body_gyroscope_mean_angle_z**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body gyroscope - angle Z

- **fast_fourier_transform_body_gyroscope_standard_deviation_angle_x**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body gyroscope - angle X

- **fast_fourier_transform_body_gyroscope_standard_deviation_angle_y**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body gyroscope - angle Y

- **fast_fourier_transform_body_gyroscope_standard_deviation_angle_z**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body gyroscope - angle Z
    
- **fast_fourier_transform_body_acceleration_magnitude_mean**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body acceleration magnitude

- **fast_fourier_transform_body_acceleration_magnitude_standard_deviation**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body acceleration magnitude

- **fast_fourier_transform_body_acceleration_jerk_signal_magnitude_mean**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body acceleration jerk signal magnitude

- **fast_fourier_transform_body_acceleration_jerk_signal_magnitude_standard_deviation**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body acceleration jerk signal magnitude

- **fast_fourier_transform_body_gyroscope_magnitude_mean**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body gyroscope magnitude

- **fast_fourier_transform_body_gyroscope_magnitude_standard_deviation**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body gyroscope magnitude

- **fast_fourier_transform_body_gyroscope_jerk_signal_magnitude_mean**

    Type: numeric<br/>
    Description: Mean of Fast Fourier Transform (FFT) body gyroscope jerk signal magnitude

- **fast_fourier_transform_body_gyroscope_jerk_signal_magnitude_standard_deviation**

    Type: numeric<br/>
    Description: Standard deviation of Fast Fourier Transform (FFT) body gyroscope jerk signal magnitude

<hr>

## Processing Steps

The following are the steps used in run_analysis.R to create the tidy data sets.


### Step 0: Download dataset and load it into R

```{r, eval=FALSE}
# Step 0.1: Download and extract dataset zip file

print("Step 0.1: Downloading dataset...")
cat("\n")

# Step 0.1.1: Clear existing R objects from the workspace
rm(list=ls())

# Step 0.1.2: Create a directory to place the dataset
data_dir <- "../data"
if (!file.exists(data_dir)) {
    dir.create(data_dir)
}

# Step 0.1.2: Download the dataset zip file and extract the contents
url <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
filename <- "project_data.zip"
file_sep <- "/"
zipfile <- paste(data_dir, filename, sep=file_sep)
method <- "curl"

download.file(url, zipfile, method)
unzip(zipfile, exdir=data_dir)
file.remove(zipfile)

# Step 0.2: Load data into R objects

print("Step 0.2: Loading data into R...")
cat("\n")

# Step 0.2.1: Load data into the workspace
filelist <- list.files(path=data_dir)
dataset_name <- filelist[1]
dataset_dir <- paste(data_dir, dataset_name, sep=file_sep)
train_data_dir <- paste(dataset_dir, "train", sep=file_sep)
test_data_dir <- paste(dataset_dir, "test", sep=file_sep)

activity_label_file <- paste(dataset_dir, "activity_labels.txt", sep=file_sep)
feature_label_file <- paste(dataset_dir, "features.txt", sep=file_sep)
train_activity_file <- paste(train_data_dir, "y_train.txt", sep=file_sep)
train_subject_file <- paste(train_data_dir, "subject_train.txt", sep=file_sep)
train_data_file <- paste(train_data_dir, "X_train.txt", sep=file_sep)
test_activity_file <- paste(test_data_dir, "y_test.txt", sep=file_sep)
test_subject_file <- paste(test_data_dir, "subject_test.txt", sep=file_sep)
test_data_file <- paste(test_data_dir, "X_test.txt", sep=file_sep)

activity_label_data <- read.table(activity_label_file)
feature_label_data <- read.table(feature_label_file)
train_activity_data <- read.table(train_activity_file)
train_subject_data <- read.table(train_subject_file)
train_data_original <- read.table(train_data_file)
test_activity_data <- read.table(test_activity_file)
test_subject_data <- read.table(test_subject_file)
test_data_original <- read.table(test_data_file)
```


### Step 1: Merge the training and the test sets to create one data set

```{r, eval=FALSE}
print("Step 1: Merging the training and test datasets...")
cat("\n")

# Step 1.1: Label train/test data with feature names

feature_names <- as.matrix(feature_label_data[2])
colnames(train_data_original) <- feature_names[,1]
colnames(test_data_original) <- feature_names[,1]

# Step 1.2: Integrate subject and activity data into train/test data
# ----------------------------------------
# Subject | Activity | ... (Features) ...
# ----------------------------------------

train_data <- cbind(train_subject_data, train_activity_data, train_data_original)
test_data <- cbind(test_subject_data, test_activity_data, test_data_original)
subject_activity_headers <- c("subject", "activity")
colnames(train_data)[1:2] <- subject_activity_headers
colnames(test_data)[1:2] <- subject_activity_headers

# Step 1.3: Merge train and test data

merged_data <- rbind(train_data, test_data)
```


### Step 2: Extract only the measurements on the mean and standard deviation

```{r, eval=FALSE}
print("Step 2: Extracting only the mean/std measurements...")
cat("\n")

# Step 2.1: Extract the names of columns to be included

colnames <- colnames(merged_data)
std_name_indices <- grep("-std()", colnames)
mean_name_indices <- grep("-mean()", colnames)
mean_freq_name_indices <- grep("-meanFreq()", colnames)
std_names <- colnames[std_name_indices]
mean_names <- colnames[mean_name_indices]
mean_freq_names <- colnames[mean_freq_name_indices]
selected_colnames <- colnames[
    (colnames %in% std_names | colnames %in% mean_names)
    &!colnames %in% mean_freq_names]
selected_colnames <- append(subject_activity_headers, selected_colnames, 2)

# Step 2.2: Extract a subset of data containing mean/std measurements

subset <- merged_data[selected_colnames]
```


### Step 3: Use descriptive activity names to name the activities in the dataset

```{r, eval=FALSE}
print("Step 3: Replacing activity IDs with activity names...")
cat("\n")

# Step 3.1: Extract activity IDs and names

activity_ids <- as.vector(activity_label_data$V1) 
activity_names <- as.vector(activity_label_data$V2)

# Step 3.2: Replace activity ID value with activity name value

for (i in 1:length(activity_ids)) {
    subset$activity[subset$activity == i] <- activity_names[i]
}
```


### Step 4: Label the dataset with descriptive variable names

```{r, eval=FALSE}
print("Step 4: Changing variable names to descriptive names...")
cat("\n")

# Step 4.1: Replace variable names with more descriptive ones

variable_names <- colnames(subset)
variable_names <- gsub("tBody", "timed_body", variable_names)
variable_names <- gsub("tGravity", "timed_gravity", variable_names)
variable_names <- gsub("fBodyBody", "fast_fourier_transform_body", variable_names)
variable_names <- gsub("fBody", "fast_fourier_transform_body", variable_names)
variable_names <- gsub("Acc", "_acceleration", variable_names)
variable_names <- gsub("Jerk", "_jerk_signal", variable_names)
variable_names <- gsub("Gyro", "_gyroscope", variable_names)
variable_names <- gsub("Mag", "_magnitude", variable_names)
variable_names <- gsub("-mean\\(\\)", "_mean", variable_names)
variable_names <- gsub("-std\\(\\)", "_standard_deviation", variable_names)
variable_names <- gsub("-X", "_angle_x", variable_names)
variable_names <- gsub("-Y", "_angle_y", variable_names)
variable_names <- gsub("-Z", "_angle_z", variable_names)
colnames(subset) <- variable_names

# Step 4.2: Save the tidy dataset

tidy_dataset_1 <- subset
output_dir <- "../output"
if (!file.exists(output_dir)) {
    dir.create(output_dir)
}
filename <- "tidy_dataset_1.txt"
output_file <- paste(output_dir, filename, sep=file_sep)
write.table(tidy_dataset_1, file=output_file, row.names=FALSE)

# Step 4.3: Save the descriptive variable names for later use in the CodeBook

markdown_dir <- "../markdown"
if (!file.exists(markdown_dir)) {
    dir.create(markdown_dir)
}
filename <- "variable_names.txt"
varname_file <- paste(output_dir, filename, sep=file_sep)
write.table(variable_names, file=varname_file, row.names=FALSE, col.names="variable_name")

print("=====> Tidy dataset 1 saved")
cat("\n")
```


### Step 5: Create a second tidy dataset with the average per activity/subject

```{r, eval=FALSE}
print("Step 5: Creating a second tidy dataset grouped by subject and activity...")
cat("\n")

# Step 5.1: Create a second tidy datset

if (!"reshape2" %in% rownames(installed.packages())) {
    install.packages("reshape2")
}
library(reshape2)

melted_data <- melt(subset, id.vars=subject_activity_headers)
aggregate_data <- dcast(melted_data,
                        subject+activity ~ variable,
                        fun.aggregate=mean)

# Step 5.2: Save the second tidy dataset

tidy_dataset_2 <- aggregate_data
filename <- "tidy_dataset_2.txt"
output_file <- paste(output_dir, filename, sep=file_sep)
write.table(tidy_dataset_2, file=output_file, row.names=FALSE)

print("=====> Tidy dataset 2 saved")
cat("\n")
```
