---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    css: faded.css
    toc: true
    number_sections: false
---


## Loading and preprocessing the data
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
data <- read.csv("../data/activity.csv")
data$date <- as.Date(as.character(data$date, "%Y-%m-%d"))
```

## What is mean total number of steps taken per day?

####**Grouping/Summarizing Number of Stpes by Date**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
library(dplyr)
num_steps_per_day <- data %>%
    group_by(date) %>%
    summarise(total_num_steps = sum(steps))
```

####**Plotting Histogram**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
library(ggplot2)
qplot(num_steps_per_day$total_num_steps, 
      geom="histogram",
      fill=I("blue"),
      col=I("lightblue"),
      main="Distribution of Number of Steps per Day",
      xlab="Number of Steps",
      ylab="Frequency")
```
```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
ggsave("../output/plot1.png", width=6, height=2)
```

####**Calculating Mean and Median Number of Stpes per Day**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
mean <- mean(num_steps_per_day$total_num_steps, na.rm=T)
median <- median(num_steps_per_day$total_num_steps, na.rm=T)
```
```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
cat("Mean total number of steps per day:")
cat(mean)
cat("Median total number of steps per day:")
cat(median)
```

## What is the average daily activity pattern?

####**Grouping/Summarizing Number of Steps by Interval**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
library(dplyr)
num_steps_per_interval <- data %>%
    group_by(interval) %>%
    summarise(average_num_steps = mean(steps, na.rm=T))
```

####**Plotting Time-Series Line Graph**
```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
png("../output/plot2.png", width=1800, height=600)
```
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
with(num_steps_per_interval, 
     plot(x=interval, 
          y=average_num_steps, 
          type="l",
          main="Average Number of Steps per Interval",
          xlab="Interval",
          ylab="Average Number of Steps",
          xaxt="n")
     )
axis(1, at=num_steps_per_interval$interval, tck=0.01)
```
```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
dev.off()
```

####**Identifying Interval with Max Average Number of Steps**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
max = max(num_steps_per_interval$average_num_steps)
max_average_interval <- 
    num_steps_per_interval[num_steps_per_interval$average_num_steps == max, ]$interval
```
```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
cat("Interval containing max average steps:")
cat(max_average_interval)
```

## Imputing missing values

####**Calculating Total Number of NAs**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
is_na <- is.na(data)
num_na_rows <- sum(is_na)
```
```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
cat("Total number of missing values: ")
cat(num_na_rows)
```

####**Creating New Dataset with NAs Imputed**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
## impute NAs with average values for corresponding interval
imputed_data <- data
for (i in 1:dim(imputed_data)[1]) {
    if (is.na(imputed_data[i, ])) {
        interval <- imputed_data[i, ]$interval
        idx <- which(num_steps_per_interval$interval == interval)
        imputed_data[i, ]$steps <- num_steps_per_interval[idx, ]$average_num_steps
    }
}
```

####**Grouping/Summarizing Number of Steps by Date**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
library(dplyr)
imputed_num_steps_per_day <- imputed_data %>%
    group_by(date) %>%
    summarise(total_num_steps = sum(steps))
```

####**Plotting Histogram**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
library(ggplot2)
qplot(imputed_num_steps_per_day$total_num_steps, 
      geom="histogram",
      fill=I("blue"),
      col=I("lightblue"),
      main="Distribution of Number of Steps per Day",
      xlab="Number of Steps",
      ylab="Frequency")
```
      
####**Calculating Mean and Median Number of Steps per Day**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
mean <- mean(imputed_num_steps_per_day$total_num_steps, na.rm=T)
median <- median(imputed_num_steps_per_day$total_num_steps, na.rm=T)
```
```{r echo=FALSE, message=FALSE, warning=FALSE, error=FALSE}
cat("Mean total number of steps per day:")
cat(mean)
cat("Median total number of steps per day:")
cat(median)
```

## Are there differences in activity patterns between weekdays and weekends?

####**Adding Weekday/Weekend Factor Variable**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
library(dplyr)
imputed_data <- imputed_data %>%
    mutate(dayofweek = weekdays(date)) %>%
    mutate(daytype = 
               ifelse(dayofweek=="Saturday" | dayofweek=="Sunday", 
                      "weekend", "weekday")) %>%
    mutate(daytype = factor(daytype, levels=c("weekday","weekend")))
```    

####**Averaging Number of Steps, Weekday vs. Weekend**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
library(dplyr)
weekday_num_steps_per_interval <- imputed_data %>%
    filter(daytype == "weekday") %>%
    group_by(interval) %>%
    summarise(daytype = "weekday",
              average_num_steps = mean(steps))

weekend_num_steps_per_interval <- imputed_data %>%
    filter(daytype == "weekend") %>%
    group_by(interval) %>%
    summarise(daytype = "weekend",
              average_num_steps = mean(steps))

merged <- rbind(weekday_num_steps_per_interval, weekend_num_steps_per_interval)
```

####**Plotting Faceted Time-Series Line Graph**
```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
library(ggplot2)
ggplot(merged, 
       aes(x=interval, y=average_num_steps)) + 
    geom_line(col="blue") + 
    facet_wrap(~daytype, ncol=1) +
    theme(strip.background = element_rect(fill="orange")) +
    ggtitle("Number of Steps per Interval Weekday vs. Weekend") +
    labs(x="Interval", y="Number of Steps") 
```    
